---
title: "NFL EDA Project"
author: "Michele Sezgin"
date: "`r Sys.Date()`"
output: openintro::lab_report
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(protoclust)
library(ggdendro)
```

### Data

```{r load-data}
# Read in the data
nfl_passing_plays <- read_csv("nfl_passing_plays_2021.csv")
# bills <- nfl_passing_plays %>%
#   filter(posteam == "BUF")
  
head(nfl_passing_plays)
```

```{r}
# Checking pass_length variable values
unique(nfl_passing_plays$pass_length)
```

### Exploratory Plots

```{r}
# Looking at distribution of downs
ggplot(nfl_passing_plays) +
  geom_histogram(aes(x = down))
```

```{r}
# Looking at distribution of play clock times when ball was snapped
ggplot(nfl_passing_plays) +
  geom_histogram(aes(x = play_clock))
```

```{r}
# Looking at time when ball was snapped vs length of yards of pass
ggplot(nfl_passing_plays) +
  geom_point(aes(x = play_clock, y = air_yards))
```

```{r}
# Visualizing pass length vs whether the pass was completed
ggplot(nfl_passing_plays, aes(x = pass_length, fill = as.factor(complete_pass))) +
  geom_bar(position = "dodge")
```


### Explanation of Data

Explanation of the data structure of the dataset

### Hypotheses

Three hypotheses you are interested in exploring

Idea: Probability of winning the game if you're winning at halftime (across a couple teams?)

-Maybe we should just focus on one team

-Pass length based on down?

-QB hit vs whether the pass was deep (player vs whole league?)
--> or QB hits vs completions (best offensive: Tom Brady, middle offensive: Ryan Tannehill, worst: Joe Burrow)
--> interceptions

```{r}
# Quarterbacks with differing line qualities
qbs <- nfl_passing_plays #%>%
  # commented out to show all quarterbacks:
  #filter(passer_player_name %in% c("T.Brady", "R.Tannehill", "J.Burrow"))

# Replacing 1s and 0s with True and False
qbs$complete_pass <- replace(qbs$complete_pass, qbs$complete_pass == 0, "False")
qbs$complete_pass <- replace(qbs$complete_pass, qbs$complete_pass == 1, "True")
qbs$qb_hit <- replace(qbs$qb_hit, qbs$qb_hit == 0, "False")
qbs$qb_hit <- replace(qbs$qb_hit, qbs$qb_hit == 1, "True")
```

``` {r}
ggplot(qbs, aes(x = qb_hit, fill = complete_pass)) +
  geom_bar(position = "dodge") +
  # commented out to show all quarterbacks
  #facet_wrap(~ passer_player_name) +
  theme_bw() +
  labs(title = "Completed Passes by QB hits", x = "QB Hit", y = "Count", fill = "Pass Completed?") #+
  #scale_fill_manual(values = c("purple", "orange"))
```

```{r}
# Visualizing whether hits and completed passes are independent
mosaicplot(table("QB Hit" = qbs$qb_hit,
                 "Completed Pass" = qbs$complete_pass),
           main = "QBs Less Likely to Complete Pass When Hit",
           shade = TRUE) # Shading by pearson residuals
```

This plot visualizes that these variables are independent 

```{r}
ggplot(qbs) +
  geom_bar(aes(x = pass_location, fill = complete_pass), position = "dodge") +
  geom_errorbar()
```


```{r}
# When the group was trying to figure out how to plot complete passes by time left in the half, I was making this function:

# nfl_passing_plays %>%
#     filter(half_seconds_remaining %in% c(1800:1700), game_half == "Half1") %>%
#     summarize(total_cp = sum(complete_pass))[[1]]

pass_completions <- function(interval, half) {
  # return total number of hits based on time interval and half
  total_cp <-  nfl_passing_plays %>%
    filter(half_seconds_remaining %in% c(interval[1]:interval[2]), game_half == half) %>%
    summarize(total_cp = sum(complete_pass))
 #return(interval[1])
  return(total_cp[[1]])
}

list_ranges <- list()
for (i in seq(1800, 100, -100)) {
  list_ranges <- append(list_ranges, c(i, i-100))
}

map(list(c(1800,1700), c(1699, 1600), c(1599, 1500), c(1499, 1400), c()), pass_completions, "Half1")

```

```{r}
# Testing out Adriana's code... it does not do what we want
nfl_passing_plays %>%
  mutate(TotalCompletions = sum(complete_pass)) %>%
  filter(game_half == "Half1") %>%
  select(TotalCompletions, game_half, half_seconds_remaining)
  
```



### Data Visualizations

Three data visualizations exploring the hypotheses, at least two of which must be multivariate. Each visualization must be in a different format from the other two, and you must have at least one categorical and one continuous visualization.

### Clustering

One clustering example

```{r}
# Trying to find separable data for clustering... this data was not separable
ggplot(nfl_passing_plays, aes(x = yards_after_catch, y = epa)) +
  geom_point()
```

```{r}
# epa vs air yards for each passing team... data looks more separable
nfl_passing_plays %>%
  filter(!is.na(air_yards)) %>%
  group_by(posteam) %>%
  summarise(epa = sum(epa), ay = sum(air_yards)) %>%
  ggplot(aes(x = epa, y = ay)) +
  geom_point()

# epa vs air yards for each passing player... data again looks more separable
nfl_passing_plays %>%
  filter(!is.na(air_yards)) %>%
  group_by(passer_player_id) %>%
  summarise(epa = sum(epa), ay = sum(air_yards)) %>%
  ggplot(aes(x = epa, y = ay)) +
  geom_point()

# epa vs yards gained for each passing player, Nick's idea... data again looks more separable
nfl_passing_plays %>%
  group_by(passer_player_id) %>%
  summarise(yg = sum(yards_gained), epa = sum(epa)) %>%
  mutate(std_yg = as.numeric(scale(yg)), std_epa = as.numeric(scale(epa))) %>%
  ggplot(aes(x = std_yg, y = std_epa)) +
  geom_point()

# Testing out how data looks if standardized first. Nick and I were debating whether to # standardize before grouping and summarizing. Prof. Yurko told us to do all wrangling  # and standardize only right before clustering
nfl_passing_plays %>%
  mutate(std_yg = as.numeric(scale(yards_gained)), std_epa = as.numeric(scale(epa))) %>%
  group_by(passer_player_id) %>%
  summarise(std_yg = sum(std_yg), std_epa = sum(std_epa)) %>%
  ggplot(aes(x = std_yg, y = std_epa)) +
  geom_point()
```

```{r}
# Minimax clustering of total epa vs yards gained for each player so we can have one    # representative player per cluster 
team_clust <- nfl_passing_plays %>%
  group_by(passer_player_id) %>%
  summarise(yg = sum(yards_gained), epa = sum(epa))

team_clust_std <- team_clust %>%
  mutate(std_yg = as.numeric(scale(yg)),
         std_epa = as.numeric(scale(epa)))

team_dist <- dist(dplyr::select(team_clust_std,
                                  std_yg, std_epa))

nfl_protoclust <- protoclust(team_dist)

ggdendrogram(nfl_protoclust, labels = FALSE, leaf_labels = FALSE,
             theme_dendro = FALSE) +
  labs(y = "Dissimilarity between clusters") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank()) 
```

```{r}
# More minimax clustering and visualizing process
minimax_player_clusters <- protocut(nfl_protoclust, k = 3)

team_clust_std %>%
  mutate(player_clusters = 
           as.factor(minimax_player_clusters$cl)) %>%
  ggplot(aes(x = std_yg, y = std_epa, color = player_clusters)) +
  geom_point(alpha = 0.5) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")

nfl_prototypes <- team_clust_std %>%
  slice(minimax_player_clusters$protos)

team_clust_std %>%
  mutate(player_clusters = 
           as.factor(minimax_player_clusters$cl)) %>%
  ggplot(aes(x = std_yg, y = std_epa, color = player_clusters)) +
  geom_point(alpha = 0.5) +
  # geom_point(data = mutate(nba_prototypes, 
  #                          player_clusters = as.factor(c(1, 2, 3))),
  #            size = 5) +
  #geom_label(data = mutate(nfl_prototypes, 
  #                         player_clusters = as.factor(c(1, 2, 3))),
  #           aes(label = passer_player_name)) +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")


# table("Clusters" = minimax_player_clusters$cl,
#       "Positions" = nba_filtered_stats$pos)
```


```{r}
# Again trying to get complete passes by time left in the half... unsuccessfully
nfl_passing_plays %>%
  filter(complete_pass == 1) %>%
  select(complete_pass, )
  ggplot(aes(x = half_seconds)) +
  geom_histogram() +
  facet_wrap(~game_half)
```



### Conclusions

Conclusions reached for the hypotheses based on your EDA and data visualizations.

...

